---
editor_options: 
  markdown: 
    wrap: 72
---

; --- title: "Clustering Lab" subtitle: "STA5077 Unsupervised Learning"
output: html_document: code_folding: hide df_print: paged ---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
source(file.path("R", "lab-helpers.R"), local = FALSE)
check_lab_packages()

dat <- load_lab_data() #read.csv("/Users/shirleyknight/Downloads/STA5077-clustering-lab/data/assignment-data.txt", sep = '', header = FALSE)
  #load_lab_data()
sapply(dat, summary)
sapply(dat, sd)
```

# 1. Initial Planning

OBJECTIVE - FIND HOW MANY CLUSTERS IN THE DATASET.

**Instruction — 10 minutes.** Agree on the purpose of the analysis, how
similarity should be defined, plausible algorithms, and the evidence
needed for a final decision.

Purpose of the analysis - find a defensible number of clusters in the
data.

Similarity to be defined - since it is continuous numerical data, and we
do not have any initial information to prompt the use of Manhattan
distance, we will use the Euclidean distance.

Plausible algorithms - K-Means and Agglomerative Hierarchical Clustering
(AHC), to evaluated the plausibility of a nested cluster structure.

Evidence needed for final decision - diagnostics (silhouette width,
elbow plot, gap statistic) and then stability testing (different
seeds/initialisation for K-means, agreement between AHC and K-means,
subsampling).

**Plan.** Record your decisions and reasons before analysing the data.

Other decisions to be made - scaling/standardising the data (evaluated
through EDA), treatment of missing values (if there are any).

**Plan discussion.** Revise the plan only where the discussion gives you
a reason.

Order of plan important to understand the dependencies. 1) Variable
Analysis - EDA, summary statistics (outliers, comparable numerical
scales, missing values)

>  

```{r planning-analysis}

```

**Analysis and decision.** Record the agreed workflow, settings and
decision criteria.

>  

**Class feedback.** Add one useful point raised by another group.

>  

# 2. EDA and Data Preparation

**Instruction — 15 minutes.** Decide what you need to learn about the
data before clustering, then perform the checks needed to support your
preparation decisions.

**Plan.** What will you inspect, and what decision will each check
inform?

We will check each feature - statistics, distribution
(scaling/transforming), missing values.

No missing values - important as we can't do Euclidean distance with it.
Outliers - summary stats & marginal distribution

Summary stats - scaling decision Scatter plot - outliers, overlap, see
something about cluster geometry (long/short/nested/orientation etc -
informs clustering method). Plot most important here because only 2
variables. Histograms - most useful for \>2 variables. Correlation
important - checking for redundant features.

**Plan discussion.** Record any justified change before coding.
Comparable ranges, although V2 has a much lower minimum. Very similar
standard deviation.

```{r eda-analysis}
# distributions
plot_marginal_distributions(dat)
plot_feature_scatter(dat)

# missing values 
any(is.na(dat))

# very low correlation
cor(dat)

dat |>
  ggplot(aes(x = V1))+ geom_boxplot() 

```

**Finding and decision.** State what you learned and any preparation you
will apply. Give a reason.

No scaling - comparable scales. Appears to be suitable for K-means and
AHC clustering. Will carry both algorithms through the analysis. No
redundant features for analysis.

**Class feedback.** Record whether the discussion changed your decision.

>  

# 3. Choosing K

**Instruction — 20 minutes.** Compute diagnostic evidence over a
defensible range of K and use it to choose a small shortlist for
sensitivity analysis.

**Plan.** Specify the K range, algorithms, fitting settings and decision
criteria before examining the results.

>  

**Plan discussion.** Record any justified change before implementation.

>  

```{r choose-k-analysis}

# calculate the objective function for each value of k 
wcss <- sapply(2:20, 
       function(k){
         kmeans(x = dat,
                centers = k,
                nstart = 25)$tot.withinss
       }
       )
# helper function needs k, the algorithm, and the objective function value for each k 
k_results <- data.frame(k = 2:20, algorithm = "K-means", objective = wcss)
plot_elbow_curve(k_results)
```

The supplied plotting functions expect your results table to contain
`k`, `algorithm`, `objective` and `avg_silhouette` columns.

```{r choose-k-plotting-helpers, eval=FALSE}
plot_elbow_curve(k_results)


d <- dist(dat)

avg_sil <- sapply(
  2:20,
  function(k){
    set.seed(5077)
    fit_k <- kmeans(x = dat, 
                centers = k, 
                nstart = 25, 
                algorithm = "Lloyd")
    mean(silhouette(fit_k$cluster, dist = d)[,3])
  }
)
  
k_results2 <- data.frame(k = 2:20, algorithm = "K-means", objective = wcss, avg_silhouette = avg_sil )
plot_silhouette_curve(k_results2)

gap <- clusGap(
  dat,
  FUNcluster = kmeans,
  nstart = 25, 
  K.max = 20,
  B = 50
)

plot(gap, main = "Gap Statistic")
maxSE(gap$Tab[,"gap"],
      gap$Tab[,"SE.sim"])
```

**Finding and decision.** Give your shortlist and explain the evidence
for retaining each candidate.

Avg silhouette width maximum at K = 15 clusters.

Gap stat 1SE rule found optimal at K = 3 clusters (local optimum). From
plot, K = 3 and K = 6 look good - simplified clusters may be more
stable. Global optimum is at K = 15. From K = 15 onwards - small
difference in avg silhouette width.

**Class feedback.** Record one alternative shortlist or interpretation.

>  

# 4. Sensitivity Analysis

**Instruction — 22 minutes.** Test whether the shortlisted solutions,
and the preferred K, are stable to initialisation and changes in the
observations analysed.

**Plan.** Choose feasible repetitions and state what would count as
concerning instability.

>  

**Plan discussion.** Record any justified change before implementation.

>  

```{r sensitivity-analysis}
# tried below, ignore if not useful
fits_3 <- lapply(
  1:100,
  function(s){
    set.seed(s)
    kmeans(dat,
           centers = 3,
           nstart = s)$cluster
  }
)
fits_6 <- lapply(
  1:100,
  function(s){
    set.seed(s)
    kmeans(dat,
           centers = 6,
           nstart = s)$cluster
  }
)
fits_15 <- lapply(
  1:100,
  function(s){
    set.seed(s)
    kmeans(dat,
           centers = 15,
           nstart = s)$cluster
  }
)

plot_initialisation_objectives()
```

Available plotting helpers, depending on the analyses your group
performs:

```{r sensitivity-plotting-helpers, eval=FALSE}
plot_initialisation_objectives(initialisation_results)
plot_subsample_mismatches(subsample_results)
plot_kstar_frequency(kstar_results)
```

**Finding and decision.** Which candidates remain credible? Distinguish
initialisation sensitivity from sensitivity to the data.

>  

**Class feedback.** Record one limitation of the sensitivity evidence.

>  

# 5. Cluster Analysis and Final Selection

**Instruction — 23 minutes.** Compare the retained partitions and
recommend one solution, revise the analysis, or conclude that no
solution is sufficiently defensible.

**Plan.** State how you will compare the candidates before viewing all
final plots.

>  

**Plan discussion.** Record any justified change before implementation.

>  

```{r final-analysis}

```

The plotting helpers take assignments and model results that you create
in the analysis block.

```{r final-plotting-helpers, eval=FALSE}
plot_candidate_clusters(
  data = dat,
  assignments = candidate_assignments,
  representatives = candidate_representatives,
  candidate_label = candidate_label
)

plot_candidate_silhouette(
  silhouette_result = candidate_silhouette,
  candidate_label = candidate_label
)
```

**Finding and final decision.** State your recommendation, supporting
evidence and most important reservation.

>  

**Class feedback.** Record how your conclusion compares with the class
decision.

>  
