---
title: "Kmeans Question Bank"
author: "Moi"
format: html
editor: visual
---

# Section A MCQ

1.  B
2.  A
3.  C
4.  A/C
5.  B
6.  A
7.  C
8.  B
9.  A
10. A
11. D
12. C
13. A/B
14. B
15. A
16. D

# Section B: Short-Answer Questions \[14 marks\]

3.1 Question B1: One k-means iteration \[3 marks\]

```{r}
K=2
u1=c(1,1)
u2=c(7,7)

A=c(1,1)
B=c(1,2)
C=c(2,1)
D=c(7,7)
E=c(8,7)
F=c(7,8)

all_points=list(A,B,C,D,E,F)

cluster_assin=c()

for(i in 1:length(all_points)){
  x1= rbind(all_points[[i]],u1)
  x2=rbind(all_points[[i]],u2)
  c1=dist(x1)
  c2=dist(x2)
  cluster_i=min(c1,c2)
  cluster_assin=c(cluster_assin,ifelse(cluster_i==c1,1,2))
  
}

cluster_assin


cl_1_mean=apply(rbind(A,B,C),2,mean)
cl_2_mean=apply(rbind(D,E,F),2,mean)
cl_1_mean;cl_2_mean
```

**3.2 Question B2: Feature scale \[2 marks\]**

In USArrests, the standard deviations of Murder, Assault, UrbanPop and Rape are approximately 4.36, 83.34, 14.47 and 9.37. The aim is to group states by an overall crime profile in which all four features should contribute comparably. Would you standardise the features before fitting k-means? Give a reason for your decision and state one reason why standardisation should not be automatic in every application.

-   Yes I would. The sd for Assualt is comparably bigger amoungst the features and would most probably artificially dominate the clustering. Thus all features would be standardised

-   Standardization is not automatic as it changes the clustering of the observations. If features are measured in the same and comparable scale, then observation values can be used directly.

3.3 Question B3: Silhouette calculation \[3 marks\]

For one observation, 𝑎(𝑖) = 1.20 and 𝑏(𝑖) = 1.50.

a\. Calculate its silhouette value. \[1 mark\]

```{r}
a_i=1.2
b_i=1.5

s_i=(b_i-a_i)/(max(c(a_i,b_i)))
s_i # close to 0, indicating that the point is balanced and equidistant from its own cluster and its nearest cluster
```

b\. Interpret the value. \[1 mark\]

-   close to 0, indicating that the point is balanced and equidistant from its own cluster and its nearest cluster

c\. State one sensible follow-up check for this observation. \[1 mark\]

Check the stability of cluster assignment by checking how this observation would be assigned by another algorithm (e.g PAM) or through subsampling.

**3.4 Question B4: Stability evidence \[2 marks\]**

At 𝐾 = 4 solution is identical across 100 seeds when nstart = 25. Agreement with PAM is 84%, and its adjusted Rand index has a low lower tail across 90% subsamples. What does this evidence say about the strengths and weaknesses of the solution? Include one caution about what stability can establish

-   84% agreement suggests there is evidence of meaningful clustering for most of the data, suggesting there is a reasonable cluster structure in the data; the 84% agreement with PAM provides evidence that the K=4 solution is reasonably consistent across two different clustering methods

-   However, the clustering does seem to have some sensitivity to the data itself, as ARI scores with subsampling show lower values, which can affect the stability of some cluster assignments at a chosen K.

-   One could check if the same points/observations are changing clusters for both stability checks. If so, it would also be worth checking the silhouette scores associated with these points

-   Expanded answer:\
    **The identical solution across 100 seeds indicates that the** K=**4 K-means solution is stable with respect to random initialization, suggesting that the solution is not strongly affected by local initialization. The 84% agreement with PAM provides further evidence that the clustering is reasonably robust to the choice of clustering algorithm. However, the low lower tail of the ARI distribution under 90% subsampling suggests that the solution has some sensitivity to perturbations in the data, with some resamples producing substantially different cluster assignments. It would therefore be useful to identify which observations are unstable and examine their silhouette widths to determine whether they are poorly separated or lie near cluster boundaries. Importantly, stability demonstrates reproducibility under the particular perturbations tested; it does not by itself establish that the clusters represent meaningful or true underlying groups.**

**3.5 Question B5**

**Decide whether either candidate should be used \[4 marks\]**

A provincial health team wants to group clinics by service-demand profile. The features were standardised because the team agreed that patient volume, waiting time, stock-out rate and travel distance should have comparable influence.

-   **Revise:** Although cluster metrics/diagnostics for K=2 are better compared to K=4 under standardized features, the conclusion/ treatment of the clinics based on the results leaves room for further analysis regarding the underlying structure of the data:

    -   the representation of the clinic- All features were standardized, so all of them had "equal" influence on the overall variation of the data

    -   Individual fit/stability: K=2 shows greater fit with higher silhouette scores and no observations with negative silhouettes, indicating that no observations were probably misclassified-The absence of negative silhouettes suggests that no clinics are clearly closer to another cluster than their assigned cluster, supporting good separation

    -    98% of observations were classified the same as PAM, whilst all observations were consistently classified in the same cluster, showing broad algorithmic agreement and stability. Also, the model is less data-sensitiv;h repeated subsampling at 10% of samples **have an ARI of at least 0.96**. ,compared to 0.55 for K=4 (more fragile model).

    -   **The 10th percentile subsample ARI of 0.96 indicates very strong stability even in the lower tail, whereas K=4's value of 0.55 indicates substantially greater sensitivity to subsampling.**

    -   However, despite the stability of the clustering at K=2, there still is no clear evidence that the clusters are practically useful given the conclusion of equal intervention treatment for the two groups. This might suggest that there may not be an actual cluster substructure in the data.

    -   To further verify/ look for evidence, a gap statistic analysis would be useful to compare the gap at different values of K against a reference null distribution to see if there is evidence the the extra clustering (K\>1) captures more information about the underlying data structure as opposed to mere noise.

    -   Also, further analysis on the distributions of the key features for the different clusters would be useful to ascertain whether certain features hold higher the average values for the different clusters. If they are broadly the same, this may be further evidence that there may not be a clear cluster structure in the data.

    -   Recheck if the features are actually meaningful to the spread of the data/clusters

# Lab Section

```{r}
lab <- read.csv("kmeans-lab-toy.csv") # load one row per service site
x_raw <- lab[c("demand_score", "complexity_score")] # keep only clustering features
# Complete each call using the inspection described by its comment.



head(lab)

```

```{r}
dim(lab) # show dimensions, variable names and data types
summary(x_raw) # show min, quartiles, median, mean and max by feature
colSums(is.na(x_raw)) # count missing observations in each feature
```

```{r}
# Plot the two features and label every site.
plot(
x_raw$demand_score, x_raw$complexity_score,
xlab = "Demand score", ylab = "Complexity score",
pch = 19
)
text(
x_raw$demand_score, x_raw$complexity_score,
labels = lab$site_id, pos = 3, cex = 0.65
)
```

```{r}

feature_sd <- sapply(x_raw,sd) # calculate the standard deviation of each feature
feature_range <- sapply(x_raw, function(x){ max(x)-min(x) }) # calculate max minus min
feature_sd # compare typical feature spreads
feature_range # compare full observed feature ranges
use_scaled <-FALSE # enter TRUE to standardise or FALSE to retain original indices
x <- if (use_scaled) scale(x_raw) else x_raw # create the chosen representation

# Featur
```

Feature scales and SDs are quite close and comparable to each other in their original scales. Standardizing the data widens this gap- thus the data in its original scale will be used

```{r}
library(cluster) # provide silhouette() and PAM for later questions
ks_wcss <- 1:5
wcss <- sapply(ks_wcss, function(k) {
set.seed(5077) # make the fit for this K reproducible
fit_k <-kmeans(x, k) # fit k-means to x with centres = k and nstart = 25
fit_k$tot.withinss # return total WCSS from fit_k
})
d <-dist(x) # calculate Euclidean dissimilarities once for all silhouette checks
ks_sil <- 2:5
avg_sil <- sapply(ks_sil, function(k) {
set.seed(5077) # make the fit for this K reproducible
fit_k <- kmeans(x, k) # fit k-means to x with centres = k and nstart = 25
mean(silhouette(fit_k$cluster, d)[, 3]) # return the mean silhouette width using fit_k and d
})
par(mfrow = c(1, 2)) # place the two diagnostic plots side by side
# plot total WCSS against K using points joined by lines
# plot average silhouette against K using points joined by lines

plot(2:5, avg_sil, type = "b",xlab = "K", ylab = "Average silhouette")
par(mfrow = c(1, 1)) # restore the one-plot layout
data.frame(K = ks_sil, average_silhouette = round(avg_sil, 3))
```

Highest avg silhouette score is achieved at K=3 or K=4

```{r}
chosen_k <- 3 # enter the preferred K from Lab Question 3
set.seed(5077) # make the final fit reproducible
km <-kmeans(x, chosen_k) # fit k-means to x using chosen_k and nstart = 25

table(km$cluster) # print the number of sites assigned to each cluster
km$centers[1,1]# print the fitted cluster centres
km$tot.withinss # print the total within-cluster sum of squares
# Plot assignments in the chosen representation.
plot(
x, col = km$cluster, pch = 19,
xlab = "Demand representation",
ylab = "Complexity representation"
)

points(km$centers[1,1], km$centers[1,2], col = "red", pch = 8, cex = 2,lwd=2) # add the fitted centres using pch = 8, cex = 2 and lwd = 2

points(km$centers[2,1], km$centers[2,2], col = "red", pch = 8, cex = 2,lwd=2) # add the fitted centres using pch = 8, cex = 2 and lwd = 2

points(km$centers[3,1], km$centers[3,2], col = "red", pch = 8, cex = 2,lwd=2) # add the fitted centres using pch = 8, cex = 2 and lwd = 2

```

```{r}
library(dplyr)
sil <- silhouette(km$cluster, d) # calculate silhouettes from km$cluster and Euclidean dissimilarities

site_ids=lab|> pull('site_id')


 # plot the silhouette values

barplot(sil[, 3], 
        names.arg = site_ids, 
        horiz = TRUE,# Add labels below bars
        col = "skyblue",            # Set fill color
        border = "black",           # Set border color
        main = "Silhouette Scores for K=3",    # Add main title
        xlab = "Store ID",        # Add X axis label
        ylab = "s(i)")            # Add Y axis label


# Continue Tomorrow

weak <- which(sil[, 3] %in% sort(sil[, 3])[1:3])# obtain row indices of the three smallest silhouette widths

cbind(
lab[weak, ], # show the site identifiers and original feature values
assigned_cluster = sil[weak, "cluster"], # show fitted cluster membership
silhouette = round(sil[weak, "sil_width"], 3) # show rounded widths
)
seeds <- 1:20 # use the same seed values in both comparisons
one_start_wcss <- sapply(seeds, function(s) {
set.seed(s) # control the initialisation for this run
km_1=kmeans(x,centers=chosen_k)# fit chosen_k with one start and return total WCSS
km_1$tot.withinss
})
many_start_wcss <- sapply(seeds, function(s) {
set.seed(s) # control the initialisations for this run
km_s=kmeans(x,centers=chosen_k,n=25) # fit chosen_k with 25 starts and return total WCSS
km_s$tot.withinss
})
# Compare the objective values produced by one and many starts.
boxplot(
list(`one start` = one_start_wcss, `25 starts` = many_start_wcss),
ylab = "Total WCSS"
)

```

One obs with a negative silhouette score (closer on average to its nearest clusters)- although at various initialization assignment of the clusters stays pretty consistent with median TWSS staying consistent with little to no variation.

```{r}
library(cluster)
library(factoextra)
pam_fit <- pam(x, k = chosen_k, metric = "euclidean")# fit PAM to x using chosen_k and Euclidean dissimilarity


comparison <-table(pam_fit$clustering, km$cluster)  # cross-tabulate km$cluster against pam_fit$clustering
comparison # print the cross-tabulation; cluster labels may be permuted
profiles <- # calculate original-scale feature means by k-means cluster
profiles # print the profiles for interpretatio
```
